# 3. Logistic regression


```{r,echo=FALSE,warning=FALSE}
#Required packages
library(tidyverse)
library(janitor)

```

## Data description

```{r}

#Read data
alc <- readr::read_csv("data/alc.csv")

names(alc)
```
**The analysis includes student attributes and average of course grades for math and portuguese exams. See full full description for the variables below:** 

1 school - student's school (binary: "GP" - Gabriel Pereira or "MS" - Mousinho da Silveira)
2 sex - student's sex (binary: "F" - female or "M" - male)
3 age - student's age (numeric: from 15 to 22)
4 address - student's home address type (binary: "U" - urban or "R" - rural)
5 famsize - family size (binary: "LE3" - less or equal to 3 or "GT3" - greater than 3)
6 Pstatus - parent's cohabitation status (binary: "T" - living together or "A" - apart)
7 Medu - mother's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
8 Fedu - father's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
9 Mjob - mother's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
10 Fjob - father's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
11 reason - reason to choose this school (nominal: close to "home", school "reputation", "course" preference or "other")
12 guardian - student's guardian (nominal: "mother", "father" or "other")
13 traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
15 failures - number of past class failures (numeric: n if 1<=n<3, else 4)
16 schoolsup - extra educational support (binary: yes or no)
17 famsup - family educational support (binary: yes or no)
18 paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
19 activities - extra-curricular activities (binary: yes or no)
20 nursery - attended nursery school (binary: yes or no)
21 higher - wants to take higher education (binary: yes or no)
22 internet - Internet access at home (binary: yes or no)
23 romantic - with a romantic relationship (binary: yes or no)
24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25 freetime - free time after school (numeric: from 1 - very low to 5 - very high)
26 goout - going out with friends (numeric: from 1 - very low to 5 - very high)
27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
29 health - current health status (numeric: from 1 - very bad to 5 - very good)
30 absences - number of school absences (numeric: from 0 to 93)

**Grades are average grades from math and portuguese exams:**
31 G1 - first period grade (numeric: from 0 to 20)
31 G2 - second period grade (numeric: from 0 to 20)
32 G3 - final grade (numeric: from 0 to 20, output target)

## Data analysis:

In this analysis I study the relationship of student attributes and alcohol consumption. The four attributes of interest studied are

- Medu - mother's education
- failures - number of past class failures
- famrel quality of family relationships
- freetime - free time after school

## Hypotheses
My hypothesis is that alcohol consumption is positively correlated with number of failures and free time after school and negatively correlated with the quality of family relationships and mother's education level. 

## Descriptives
Let's check some cross tabulations and graphs to study the relationships

### Mother's education
```{r}

ggplot(data=alc, aes(x=factor(Medu),y=alc_use))+
  geom_boxplot()+
  geom_hline(yintercept = 2,linetype="dashed", color="pink")
  labs(y="Alcohol usage", x="Mother's education", title = "Alcohol consumption by mothers education")

```
From the picture above we see that we see that average consumption for all education levels is between 1 and 2.5. Only in category 3 that is "secondary education" the average consumption is considered as high use. No clear pattern can be observed from this. No education mean seems to be sligtly higher than for others. But when we cross tabulate and see how many student have mother in category 1 we see that there is only 3 of these. Hence we should not make any conclusions based on this. 


```{r}
tabyl(alc$Medu)
```
The figure doesn't support my hypothesis that mothers education would decrease drinking. 


### Failures

```{r}
ggplot(data=alc, aes(x=factor(failures),y=alc_use))+
  geom_boxplot()+
  geom_hline(yintercept = 2,linetype="dashed", color="pink")+
  labs(y="Alcohol usage", x="failures", title = "Alcohol consumption by mothers education")+
  theme_minimal()
```

The figure above supports the hypothesis that failures is positively correlated with alcohol usage. Nearly all of the student in failure category 3 or more use alcohol more than the limit for high usage. But again after checking the number of students in that category we find that they are very rare cases. 


```{r}
tabyl(alc$failures)
```


### Quality of family relationships


Check first the distribution of student to the categories. No using for visuals instead of cross tabulations:

```{r}
ggplot(data=alc, aes(x=famrel))+
  geom_histogram(fill="pink") +
  labs(y="Number of students", x=" 1-very bad to 5 - excellent", title = "Students by quality of family relationships")+ 
  theme_minimal()
```

We see that most of the students have quite good family relationships. Lets then check how alcohol usage is distributed in each of these categories

```{r}
ggplot(data=alc, aes(x=alc_use))+
  geom_histogram()+
  geom_vline(xintercept = 2,linetype="dashed", color="pink")+
  labs(y="Number of studens", x="alchol consumption", title = "Alcohol consumption by mothers education")+
  facet_grid(~famrel)
```

Above we see that on the contrary to my hypothesis it doesn't seem that alcohol consumption would be higher when the quality of family relationships are lower. At least in the sample most of students using a lot are in the panes 3 to 5. 


### Free-time
Finally lets check free time after school and alcohol consumption. 

```{r}

ggplot(data=alc, aes(x=freetime))+
  geom_histogram(fill="pink") +
  labs(y="Number of students", x=" 1-very low to 5 - high", title = "Students by quality of family relationships")+ 
  theme_minimal()

```


Bulk of the students are in the middle of the scale.


```{r}
ggplot(data=alc, aes(x=factor(freetime),y=alc_use))+
  geom_boxplot()+
  geom_hline(yintercept = 2,linetype="dashed", color="pink")+
  labs(y="Alcohol usage", x="Free time", title = "Alcohol consumption by after school free time")+
  theme_minimal()
```

From the above we see that alcohol consumption is on average higher for students with lots of after school free time. Perhaps students with low free time spend all their time at sports or similar activities and don't simply have time to drink. 


## Logistic regression

### Estimating the model
```{r}

model <- glm(high_use ~ Medu + failures + famrel + freetime ,family=binomial(link='logit'),data=alc)
summary(model)
```

### Prediction vs. actual 
```{r}
alc <- mutate(alc, probability = predict(model, type = "response"))
alc <- mutate(alc, prediction = probability > 0.5)


tabyl(alc, high_use,prediction)
```

It seems that model does pretty good job of predicting non-high-users: 259 non-high-users the models predicts 244 correctly. On the other hand the model is terrible predicting the high-users. From 111 high users in the sample it predicts correctly only 16. 

### The total proportion of inaccurately classified individuals

```{r}
inaccurate <- nrow(alc %>% filter(high_use!=prediction))
all <- nrow(alc)

inaccurate/all
```
The total proportion is of inaccurately classified individuals is around 30 percent. This is definitely improvement for a single coin toss strategy. 

###Bonus: 10-fold cross validation



