# 2. Regression and model validation

```{r, warning=FALSE}
library(tidyverse)
library(sandwich)
library(lmtest)
```

## Read the data

```{r}
analysis_data <- read_csv("data/learning2014.csv")


str(analysis_data)

```

Data set includes survey data from students, source Kimmo Vehkalahti, extra info [here](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-meta.txt)

-   Attitude is a sum of 10 questions related to students attitude towards statistics, each measured on the Likert scale (1-5)
-   deep is a mean of 11 questions related to deep learning
-   stra is a mean of 8 questions related to strategic learning
-   surf is a mean of 10 questions related to surface learning - points are exam point (observations with points equal to zero are dropped)

## Graphical overview:

```{r}

data_long <- analysis_data %>% select(attitude,deep,stra,surf) %>% 
  pivot_longer(everything(), names_to = "Variable")

ggplot(data_long, aes(x = Variable, y = value)) +
  geom_boxplot() +
  labs(title = "Figure 1: Distribution for four questions categories",
       x = "Question category",
       y = "Value") +
  theme_minimal()


```

Figure one plots the distributions of attitude, deep, stra and surf questions. Median value varies across roughly 2.75 to 3.75. Only deep and surf categories have some outlier observations.

```{r}
ggplot(data=analysis_data,aes(x=Age, fill=gender))+
  geom_histogram(bins = 20) +
  labs(title = "Figure 2: Age Distribution by Gender", x = "Age", y = "Frequency") +
  theme_minimal()

```

Figure 2 plots age distribution of participants. Around 50 percent of participants are around 20 yrs old. Most participants females and some outliers, mainly men.

```{r}
ggplot(analysis_data, aes(x = gender, y = Points)) +
  geom_boxplot()+
  theme_minimal()+
  labs(title = "Figure 3: Points distribution by Gender", x = "Gender", y = "Points") 
```

Figure show the distribution of points by gender. Based on this it seems that the points are distributed quite similarly acrross to gender types in the data.

## Regression model

```{r}
#Run model
lm1 <- lm(formula= Points ~ attitude + deep  + Age  + surf, data=analysis_data)
summary(lm1)

```

Based on outputs above one can conclude that only statistically significant coefficient in the specification is attitude (assuming homoskedastic errors!). On average one-unit increase in the scale of attitude increases points by roughly 3.5. 

-  t-value tests whether the coefficient is statistically significantly different from zero
-  F-statistics tests whether all of the coefficient are different from zero (jointly)

```{r}
lm2 <- lm(formula= Points ~ attitude, data=analysis_data)
summary(lm2)

```

After excluding non-significant variables we see that is still aroung 3.5. Multiple R-squared measures the proportion of variance in the dependent variable that can be explained by the predictor variables in the model. It ranges from 0 to 1, with higher values indicating a better fit. In the output above we see that this is 0.1906 so around 20 percent. 

## Diagnostic plots

```{r}
# Residuals vs Fitted values
plot(lm2, which = 1,caption = "Figure 3: Residuals vs fitted")
```

```{r}
qqnorm(lm2$residuals)
qqline(lm2$residuals)
```

```{r}
plot(lm2, which = 5)

```

Model assumptions:

1. Linear relationship between predictors and outcome
2. Independence of residuals
3. Normal distribution of residuals
4. Equal variance of residuals

Validity of our assumptions:

-  Based on Figure 3. the assumption on linear relationship could seem plausible
- Based on Q-Q plot we see that residuals are diverging from the line, which suggest that errors are not normally distributed
- Based on Figure 3. the assumption on equal variance seems plausible, no clear patterns that suggested otherwise.


